NumPy---Array Programming in Python

### 2020-02-14 body.tex
Flesh out bullet points in discussion section

NumPy combines the expressive power of \emph{array programming},
the performance of C, and
the readability, usability, and versatility of Python in a mature,
well-tested, well-documented, and community-developed library.
Many fast algorithms are already implemented by libraries in the
scientific Python ecosystem.
Where extreme optimization is warranted, compiled languages such as
Cython \cite{behnel2011cython}, Numba \cite{Lam:2015:NLP:2833157.2833162},
and Pythran \cite{guelton2015pythran}, that
extend Python and transparently accelerate bottlenecks, can be
used.
Because of NumPy's simple memory model, it is easy to write low-level, hand-optimized code, usually in C
or Fortran, to manipulate NumPy arrays and pass them back to
Python.
Furthermore, using array protocols, it is possible to utilize the full
spectrum of specialized hardware acceleration with minimal changes to
existing code.

@@

What began as an attempt to add an array object to Python became
the foundation of a vibrant ecosystem of tools.
Now, a large amount of scientific work depends on NumPy being correct,
fast, and stable.
It is no longer a small community project, but is core scientific
infrastructure.
The developer culture has changed as well.
- Governance
- Process (NEP, roadmap, how we handle issues on GH, etc.)
- still primarily draw from graduate students and researchers
- Funding (some recent evidence that that is improving  grants, or
Moore/Sloan/CZI) CZI EOSS
- Major refactoring of code
- Redesigning website and documentation to be more friendly, easy to
navigate

NumPy's is no longer just the foundational array library underlying
the scientific Python ecosystem, but has also become the standard API
for tensor computation and a central coordinating mechanism between
array types and technologies in Python.

- Protocols: will be extending that work

Over the next decade, we will face new challenges.
- AI / data science mature
- hardware / Moore's law, specialized devices
- more users (diversity + numbers)
- scale / type of data; scientific data, lightsheet microscopes, LSST
- new ideas come from other languages like Julia, Rust, ...; JIT improvements

NumPy is poised to tackle these challenges, and florish in the next decade.
Need
- Next generation of developers (grad students) to sign on
- Funders and industry continue funding this project as well as increasing support for the ecosystem through new funding mechanisms.

>>>>

NumPy combines the expressive power of \emph{array programming},
the performance of C, and
the readability, usability, and versatility of Python in a mature,
well-tested, well-documented, and community-developed library.
Many fast algorithms are implemented by libraries in the
scientific Python ecosystem.
Where extreme optimization is warranted, compiled languages such as
Cython \cite{behnel2011cython}, Numba \cite{Lam:2015:NLP:2833157.2833162},
and Pythran \cite{guelton2015pythran}, that
extend Python and transparently accelerate bottlenecks, can be
used.
Because of NumPy's simple memory model, it is easy to write low-level, hand-optimized code, usually in C
or Fortran, to manipulate NumPy arrays and pass them back to
Python.
Furthermore, using array protocols, it is possible to utilize the full
spectrum of specialized hardware acceleration with minimal changes to
existing code.

@@

What began as an attempt to add an array object to Python became the
foundation of a vibrant ecosystem of tools.  Now, a large amount of
scientific work depends on NumPy being correct, fast, and stable.  It
is no longer a small community project, but is core scientific
infrastructure.

The developer culture has matured: while initial development was
highly informal, NumPy now has a roadmap and a process for proposing
and discussing large changes.  The project has formal governance
structures and is fiscally sponsored by NumFOCUS, a nonprofit that
promotes open practices in research, data, and scientific computing.
Over the past few years, the project attracted its first funded
development, sponsored by the Moore and Sloan Foundations, and
received an award as part of the Chan Zuckerberg Initiative's
Essentials of Open Source Software program.  With this funding, the
project was (and is) able to have sustained focus over multiple months
to implement substantial new features and improvements.  That said, it
still depends heavily on contributions made by graduate students and
researchers.

NumPy's is no longer \emph{just} the foundational array library underlying
the scientific Python ecosystem, but has also become the standard API
for tensor computation and a central coordinating mechanism between
array types and technologies in Python. Work continues to expand on and
improve these interoperability features.

Over the next decade, we will face several new challenges.
%Artificial
%Intelligence and Data Science will mature ... [what do we say about
%  this?].
New devices will be developed, and existing specialized
hardware will evolve, to meet diminishing returns on Moore's law.
There will be more, and a wider variety of, data science
practitioners, a significant proportion of whom will be using NumPy.
The scale of scientific data gathering will continue to expand, with
the adoption of devices and instruments such as light sheet
microscopes and the LSST.  New generation languages, interpreters, and compilers,
such as Rust, Julia, and LLVM, will invent and determine the viability
of new concepts and data structures.

Through various mechanisms described in this paper, NumPy is poised to
embrace such a changing landscape, and to continue playing a leading
role in interactive scientific computation.  To do so will require
sustained funding from government, academia, and industry.  But,
importantly, it will also need a new generation of graduate students
and other developers to engage, to build a NumPy that meets the needs
of the next decade of data science.
